### 2009-05-21 response-to-reviews2.txt

Thank you for the detailed recommendations.  We now state clearly that
we take the phylogeny as given, although we also note in the
Discussion that in principle we could construct a basic sampler
over trees using our method (though we have not constructed or
evaluated such a sampler).

We have added more discussion of the potential rate-limiting steps and
corresponding constraints to the manuscript ("Overview").  As the
reviewer points out, perhaps the most realistic pieces of outside
information to have are a consensus structure for sequences and/or a
curated multiple alignment.  In principle these can all be
incorporated into our inference algorithms with the alignment
constraints used by, e.g., QRNA and PFOLD, or the simultaneous fold and
alignment constraints used by, e.g., Stemloc and CONSAN.

@@

> Perhaps most glaringly, the section entitled "dynamic programming 
> inference algorithms" never actually states an algorithm! It should. 
> The "loopy DP" algorithm is claimed in the abstract (and in the 
> authors' response to reviews) to be a key result of the paper, but it 
> is only referred to ("See Text S1 for more [sic] details of this 
> algorithm"). Again this is a misuse of supplementary material. A paper 
> is not supposed to be a mere advertisement of cool things we're 
> working on, and cooler things we're thinking about working on. Too 
> much of this manuscript still reads this way (though it's greatly 
> improved from the previous version). A paper should deliver crisp, 
> hard-nosed, reproducible results to a reader. In the case of an 
> algorithms paper, the reader should be able to understand and 
> implement the key ideas from a careful and complete reading of the 
> paper. Generally speaking, the authors are still spending way too much 
> time talking about ideas for their future, and too little
> teaching the reader something concrete in the present. Relegating 
> important stuff to supplementary material while spending so much room 
> on vague conceptual stuff in the text is aggravating (I would be 
> sympathetic if the manuscript were closely on point, and stuff was 
> being moved to supp material solely to keep length down and maintain 
> clarity of the essential results description). Supplementary material 
> should be used for stuff that doesn't fit (large data sets, source 
> code), is better distributed in electronic form (video, formatted 
> data) or is truly parenthetical to the main results (details that the 
> reader doesn't need to see to understand the paper). Only in a last 
> resort (in space-limited journals, which does not include PLoS 
> Computational Biology) should it contain any actual results of the paper.

Thank you for the feedback; we have attempted to revise the paper
accordingly such that it is completely self-contained and
reproducible.  We have given complete specifications of the CYK,
Inside, Outside, and CYK traceback algorithms for three sequences,
constrained by a pre-determined set of folds for each sequence ("fold
envelopes").  They are written in pseudocode which corresponds closely
to our actual implementation in the program (INDIEGRAM) which we
describe.

@@

We agree that these reconstructions were not very informative.
Unfortunately these were the only sequences that we could find which
were short enough to be tractable on the hardware we had available. We
now report simulation results instead, as suggested.

We thank the reviewer for pointing out errors in the original
figures. These were due to mistakes made during manual conversion of
indiegram's output parse trees into Stockholm-format alignments and a
bug in the code used for maximum-likelihood reconstruction of
sequences (the xrate program). We have now automated this process and
fixed the bug, verifying that this has eliminated the errors.

We removed the tRNA and nanos reconstructions and posted (corrected)
versions online at the following URL (which is linked from the
indiegram webpage),
http://biowiki.org/IndiegramExamples,
although we note in the text that the reconstructions were too easy
because of short branch lengths.

The paper still includes the runtime estimates for different length
classes of RNA genes.

@@

> More minor comments follow.
>
> title: A paper should deliver something akin to what its title seems 
> to promise. There is no result in this paper that bears meaningfully 
> on "primordial ribozymes" or "ancient RNA". The main results are 
> instead a promising but rough and very preliminary theoretical 
> framework for modeling the evolution of structural RNAs, with some 
> small toy examples of applications to three small RNA sequences. The 
> grand idea of paleogenetics motivates the work, and certainly should 
> be discussed, but it doesn't belong in the title unless you're 
> actually doing (and testing!) paleogenetics. A less grand title, more 
> consistent with results on toy three-sequence data sets, is appropriate.

We have accordingly changed the title to "Evolutionary triplet models of
structured RNA"; we welcome title suggestions from the reviewer!

@@

> p.21 in general the whole "reconstructing RNAs with a simple model" 
> section is too chatty and imprecise. "Both predicted ancestral 
> structures closely match the RFAM-annotated consensus structures", for 
> example, is not a quantitative result, and was not set up to address 
> any particular focused question (like, for instance, does our method 
> do better than some simpler approach of addressing the same problem?) 
> Again too much space is devoted to vague forward-looking conceptual 
> claims ("Though it's beyond the scope of the paper, a revised pairwise 
> model could readily be designed...") and too little attention is given 
> to the actual model and problems at hand.

We have reworded this section per our new focus on the simulation results.

> Table 2: it's unclear why we need separate states L and I_L, or S and 
> I_S; these nonterminals have identical production rules.

We distinguished between these states to help clarify our method,
where states are classified as "Start" or "Insert" to simply the
statement of the composition algorithm.  For implementation purposes
we certainly agree that it is a mistake to have "duplicate" states and
when possible, we remove these from the resulting 3-sequence SCFG.

> Similarly, I 
> don't see why we need B_e; S and I_S can produce L directly; and L and 
> I_L can produce an LS bifurcation directly, rather than invoking 
> nonterminal B.

As with L and I_L, we drew this distinction for clarity of approach,
so that we could distinguish between states which emitted nonterminals
and bifurcation states to make the composition algorithm more
intuitive.  A transition S -> B_e -> L is classified as an "empty
bifurcation" as discussed in the revised text and replaced with an
"equivalent direct transition" of precisely the form which you suggest.

> The L and I_L productions don't appear to be correctly 
> normalized, unless the (undefined) term p_S is 0.0.

The equilibrium distribution of unpaired nucleotides is co-normalized
with the stem probability p_S (labeled as \pi_l (S) for clarity in the
revised manuscript), so that the production rules for L and I_L are
properly normalized.  We have added a note to this effect in the
caption for Table 2 in the revised manuscript.

@@

> A production I_L 
> \rightarrow x I_L does not correspond to a pair-HMM rule, as the 
> caption states; this is just a standard regular grammar production.

Thank you for pointing this out.  We have corrected this in the 
revised text.

> Table 4: perhaps I'm slow, but this doesn't seem like the clearest 
> explanation of a branch transducer. I'm expecting to see rules 
> corresponding to, for example, the ancestor produced G-C from a 
> basepair state; the branch transducer chooses a basepair->basepair 
> substitution transduction as its state path with some probability, 
> then eats the GC and emits an xy. The production rules in Table 4 look 
> more like a single-sequence SCFG to me (x,y do appear in the wait 
> state conditionals, but where are x,y coming from on the left hand 
> side -- where are they being absorbed from?)
> The productions also 
> don't seem to correspond entirely to the state definitions in Table 5; 
> for example, if M_S is a "match" state describing "base pair 
> substitution" that "absorbs (x,y)" and "emits (u,v)" according to 
> Table 5, then why does the Table 4 give no production rule for 
> emitting (u,v) given (x,y)? Instead it's the wait state W_S in Table 4 
> that has the expected (x,y) -> (u,v) production rule. Maybe the
> rationale would become clear to me if I spend more time staring at 
> this, but it seems to need more explanation and/or clarification and 
> cleanup.

@@

We wrote the rules here in the Mealy-machine framework, where inputs
and outputs are attached to transitions (c.f., the Moore-machine framework,
where ouputs depend only on the current state of the machine).  The
rule
W_S -> u M_s v ;       alpha_2 (t) * p (u, v| x, y)
is intended to correspond to a transition in a Mealy machine
from state W_S to M_S, where the transition has an input of (x, y)
and an output of (u, v).  We have attempted to clarify this in the
revised text.

We apologize for the confusion; we have attempted to follow the
standard SCFG notation as closely as possible in an attempt to be
clear, but we certainly welcome further suggestions to clarify
the model.

> Figure 1: 5' and 3' ends of RNA structures should be labeled.

The 5' and 3' ends of structures are now labeled (e.g., Figures 3 and
4 of the revised manuscript).

### 2009-05-21 paper.tex
Final edits.

The statistics of the TKFST model on the three-taxon phylogeny
illustrate the advantages of our procedure for model
construction.  While the singlet and branch transducers are relatively simple---the singlet transducer,
shown in \tabref{tkfstsinglet}, has 7 total states and 2 bifurcation states and
the branch transducer, shown in \tabref{tkfstbranch}, has 21 total states
and 6 bifurcation states---the ensemble model of three extant sequences 
is very complex.
The naive exponential upper-bound gives a maximal state space of size $O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.
Using our uninformed search algorithm, we determined that there are 287 accessible states
and 686 possible transitions between these states (compare with the $287^2 \simeq 8 \cdot 10^4$ transitions
estimated with the exponential calculation).  After performing the
transformations described in ``Eliminating null cycles in SCFGs'' to
eliminate useless windback states,
the ensemble model has a reduced state space with 230 states,
albeit at the cost of extra transitions, bring the total to 1,789 transitions
(here we are trading reduced memory complexity, which is linear in the number
of states, for increased time complexity, which is linear in the number of transitions).
Note that both before and after the reduction in complexity,
the total number of states and transitions are less than the approximate bounds
of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot 3) \Rightarrow 441$ states
and
$O(a \cdot b^2 \cdot N) = O(7 \cdot 21^2 \cdot 3) \Rightarrow 9,261$ transitions
suggested in ``The composition algorithm.''

% to do: dump output of tripletscfg.pl --tkfst into supplement
Text S1 contains the complete model specification, including states and transitions,
produced by our model composition algorithm.  The extreme complexity of the
ensemble model, despite the simplicity of the underlying model of RNA
structure, makes clear the necessity for automated procedures for
model construction.

@@

% Use table to allow the table to span the width of the entire page.
\begin{table}[!ht]
  \caption{
    \textbf{Singlet transducer (single-sequence SCFG) of the TKF Structure Tree model.}}
  \begin{tabular}{|rcl|r||rcl|r|}
    \hline
    Source & $\rightarrow$ & Destination & probability & Source & $\rightarrow$ & Destination & probability \\ \hline
    $L$ & $\rightarrow$ & $u \,\, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $S$ & $\rightarrow$ & $u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$ \\ \hline
    & $|$ & $B$ & $\kappa_l \cdot \pi_l(S)$ & & $|$ & $B_e$ & $1-\kappa_s$ \\ \hline
    & $|$ & $\Send$ & $1-\kappa_l$ & & & & \\ \hline
    $I_L$ & $\rightarrow$ & $u \, \, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $I_S$ & $\rightarrow$ & $u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$\\ \hline
    & $|$ & $B$ & $\kappa_l \cdot \pi_l(S)$ & & $|$ & $B_e$ & $1-\kappa_s$\\ \hline
    & $|$ & $\Send$ & $1-\kappa_l$ & & & & \\ \hline \hline
    $B$ & $\rightarrow$ & $(L \,\, S)$ & 1 & & & & \\ \hline
    $B_e$ & $\rightarrow$ & $(L \,\, \Send)$ & 1 & & & & \\ \hline
  \end{tabular}
  \begin{flushleft}
    The state types for this model are shown in \tabref{tkfstsinglet-types}.
    The singlet transducer generates ancestral RNA sequences and
    structures.
    We use the notation of formal grammars to represent state
    transformation rules; for example, the rule $I_L \rightarrow
    u\,\,I_L$ corresponds to (in a Pair HMM) an $\Sinsert$ state $I_L$
    emitting a nucleotide $u$ and then making a self-transition.
    Both loop ($L$ and $I_L$) and stem ($S$ and $I_S$) sequence evolve as TKF sequences
    with length parameters $\kappa_l$ and $\kappa_s$ (defined in ``A simple model of RNA structural evolution'').
    $\pi_l(u)$ and $\pi_s(uv)$ are the equilibrium distributions of unpaired nucleotides $x$ and
    paired nucleotides $(u,v)$.
    The bifurcation state $B_e$ is used to end stem sequences
    (only loop sequences are allowed to transition to the empty string).
  \end{flushleft}
  \tablabel{tkfstsinglet}
\end{table}

>>>>

The statistics of the TKFST model on the three-taxon phylogeny
illustrate the advantages of our procedure for model construction.
While the singlet and branch transducers are relatively simple---the
singlet transducer, shown in \tabref{tkfstsinglet}, has 7 total states
and 2 bifurcation states and the branch transducer, shown in
\tabref{tkfstbranch}, has 21 total states and 6 bifurcation
states---the ensemble model of three extant sequences is very complex.
The naive exponential upper-bound gives a maximal state space of size
$O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.  Using our
uninformed search algorithm, we determined that there are 287
accessible states and 686 possible transitions between these states
(compare with the $287^2 \simeq 8 \cdot 10^4$ transitions estimated
with the exponential calculation).  After performing the
transformations described in ``Eliminating null cycles in SCFGs'' to
eliminate useless windback states, the ensemble model has a reduced
state space with 230 states, albeit at the cost of extra transitions,
bring the total to 1,789 transitions (here we are trading reduced
memory complexity, which is linear in the number of states, for
increased time complexity, which is linear in the number of
transitions).  Note that both before and after the reduction in
complexity, the total number of states and transitions are less than
the approximate bounds of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot
3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in ``The
composition algorithm.''  Nonetheless, the extreme complexity of the
ensemble model, despite the simplicity of the underlying model of RNA
structure, makes clear the necessity for automated procedures for
model construction.  Dataset S1 gives the state space of the composed
model constructed by the search algorithm and Dataset S2 the reduced
model after eliminating ``windback'' states; both are in Graphviz format for
visualization and show the state of the singlet transducer generating
ancestral sequence as well as the states of the branch transducers
generating observed sequences.

@@

% Use table to allow the table to span the width of the entire page.
\begin{table}[!ht]
  \caption{
    \textbf{Singlet transducer (single-sequence SCFG) of the TKF Structure Tree model.}}
  \begin{tabular}{|rcl|r||rcl|r|}
    \hline
    Source & $\rightarrow$ & Destination & probability & Source & $\rightarrow$ & Destination & probability \\ \hline
    $L$ & $\rightarrow$ & $u \,\, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $S$ & $\rightarrow$ & $u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$ \\ \hline
    & $|$ & $B$ & $\kappa_l \cdot \pi_l(S)$ & & $|$ & $B_e$ & $1-\kappa_s$ \\ \hline
    & $|$ & $\Send$ & $1-\kappa_l$ & & & & \\ \hline
    $I_L$ & $\rightarrow$ & $u \, \, I_L$ & $\kappa_l \cdot \pi_l(u)$ & $I_S$ & $\rightarrow$ & $u \,\, I_S \,\, v$ & $\kappa_s \cdot \pi_s(uv)$\\ \hline
    & $|$ & $B$ & $\kappa_l \cdot \pi_l(S)$ & & $|$ & $B_e$ & $1-\kappa_s$\\ \hline
    & $|$ & $\Send$ & $1-\kappa_l$ & & & & \\ \hline \hline
    $B$ & $\rightarrow$ & $(L \,\, S)$ & 1 & & & & \\ \hline
    $B_e$ & $\rightarrow$ & $(L \,\, \Send)$ & 1 & & & & \\ \hline
  \end{tabular}
  \begin{flushleft}
    The state types for this model are shown in
    \tabref{tkfstsinglet-types}.  The singlet transducer generates
    ancestral RNA sequences and structures.  We use the notation of
    formal grammars to represent state transformation rules; for
    example, the rule $I_L \rightarrow u\,\,I_L$ corresponds to (in a
    Pair HMM) an $\Sinsert$ state $I_L$ emitting a nucleotide $u$ and
    then making a self-transition.  Both loop ($L$ and $I_L$) and stem
    ($S$ and $I_S$) sequence evolve as TKF sequences with length
    parameters $\kappa_l$ and $\kappa_s$ (defined in ``A simple model
    of RNA structural evolution'').  $\pi_l(u)$ and $\pi_s(uv)$ are
    the equilibrium distributions of unpaired nucleotides $u$ and
    paired nucleotides $(u,v)$ and are normalized such that
    $\pi_l(S) + \sum_u \pi_l(u) = 1$ and $\sum_{u,v} \pi_s(uv) = 1$.  The bifurcation state $B_e$ is used
    to end stem sequences (only loop sequences are allowed to
    transition to the empty string).
  \end{flushleft}
  \tablabel{tkfstsinglet}
\end{table}
